---
title: "Sex Labeling Example"
author: "E Flynn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, echo=FALSE, message=FALSE}
require('exprsex')
```



## Download a dataset and prepare for sex labeling
You can download the dataset with whatever method you prefer. In the end, you need a vector for gene-to-probe mapping and the expression matrix.
### Option 1: Download with MetaIntegrator
```{r, warning=FALSE, message=FALSE}
require('MetaIntegrator')
gse <- getGEOData("GSE55668")

# grab sex labels
sex_lab <- gse$originalData$GSE55668$pheno$`Sex:ch1`

# View the expression matrix and some of the key mappings
head(gse$originalData$GSE55668$expr[,1:5])
head(gse$originalData$GSE55668$keys[!is.na(gse$originalData$GSE55668$keys)]) 
```

The function `prepInput` is then used to summarize the probes to genes and convert to ranks.
```{r}
exp_rank <- prepInput(gse$originalData$GSE55668$expr, gse$originalData$GSE55668$keys)
head(exp_rank[,1:5])
```

### Option 2: Download with GEOquery
When using GEOquery, we just have to run `createGSEObj` to convert the data.
```{r, warning=FALSE, message=FALSE}
require('GEOquery')
gse2 <- getGEO("GSE55668")

gse.obj2 <- createGSEObj(gse2)
gse.obj2$sex_lab <- pData(gse2$GSE55668_series_matrix.txt.gz)$`Sex:ch1`

```

The next steps are all the same:
```{r}
gse.obj2 <- prepInput(gse.obj2) # list_genes?
fit_g2 <- trainSexLab(gse.obj2$rank, gse.obj2$sex_lab)
fit_g3predSexLab(fit_g2, gse.obj2$rank)
```


## Reformat sex labels
The sex labels in the study are present as Female/Male - we want them as 0 (f) and 1 (m), so they must be mapped.
The end result is a vector with sex labels and the names of the vector are the sample IDs.
```{r}
head(sex_lab)
map_list <- list("Female"=0, "Male"=1)
sex_lab_reform <- unlist(sapply(sex_lab, function(x) map_list[[x]]))
names(sex_lab_reform) <- colnames(exp_rank)
head(sex_lab_reform)
```

## Train a model using the provided data and models
Fits include a sex of female-specific genes, male-specific genes, and a cutoff threshold. 
The `trainSexLab` function uses the training data and labels to remove low-variance genes from a 
```{r, echo=FALSE, message=FALSE}
fit_ex <- trainSexLab(exp_rank, sex_lab_reform)

# take a look at the fit
str(fit_ex)
```

## Predict sex labels using an existing fit
Use a general t and run sex labeling.
```{r}
data(fit_all_train)
# load and view a fit
str(fit_all_train)

# predict using existing fit
sex_lab_f <- predSexLab(fit_all_train, exp_rank)
head(sex_lab_f)
```

Compare to the sex labels from the trained fit and the actual sex labels.
```{r}
# predict using the new fit (warning: this fit was trained on the data!)
sex_lab_t <- predSexLab(fit_ex, exp_rank)
table(sex_lab_f==sex_lab_reform)
table(sex_lab_t==sex_lab_reform)
```


## Visualize the sex labels
```{r, message=FALSE, warning=FALSE}
# here we are grabbing the predicted sex labels
visualizeXYplot(expData, gene_symbol_list, sex_lab_f)
```


